---
title: "Getting started with GGPlot2"
author: "Christophe Ladroue"
date: "EdinbR 18/03/2015"
output:
  ioslides_presentation:
    fig_height: 3
    fig_width: 5
    incremental: no
    self_contained: yes
    transition: 0.2
---
```{r, echo=FALSE, message=FALSE}
library(ggplot2)
```

## Why learn GGPlot2?

- Prettier plots
- A systematic way of manipulating data, esp. when combined with `dplyr`. Better than relying on a collection of recipes.

## GGplot2:

- GG stands for Grammar of Graphics (Wilkinson, 2005).
- Provides a unified, statistically-oriented approach to graphical representation.
- Does so thanks to the use of more abstract objects than just raw graphical elements.


## Using `ggplot2`
`ggplot2` offers at least three ways for producing plots, varying in verbosity:

- `qplot`
- `ggplot+geom_xxx`
- `ggplot+layer` 

Today, we'll focus on `ggplot+geom` because it's verbose enough to make the workings of `ggplot2` transparent and readable.

## A first example 1/2

- Dataset to use: `diamonds`. `?diamonds` gives a short description of the data.
- We'll use a subset of the whole dataset:
```{r, echo=TRUE}
    set.seed(42)
    small <- diamonds[sample(nrow(diamonds),1000),] 
```
- A quick peek at the data:
```{r, echo=FALSE}
head(small)
```

## A first example 2/2

A first look at the relation between `carat` and `price`:

```{r, echo=TRUE, fig.align='center'}
ggplot(small) + 
  geom_point(aes(x=carat,y=price,colour=cut)) +
    scale_y_log10() + 
			facet_wrap(~cut) +
				ggtitle("First example")
```


## Let's decompose the `ggplot2` command. 
### Data
Create the ggplot object and populate it with data (always a data frame)
```{r, eval=FALSE}
ggplot(small)
```

### Layers
Add layer(s)
```{r, eval=FALSE}
geom_point(aes(x=carat,y=price,colour=cut))
```
## Let's decompose the `ggplot2` command. 
### [facets]
Used for conditionning on variable(s)
```{r, eval=FALSE}
facet_wrap(~cut)
```

### [scales]
Scales for dimensions, colour palettes etc.
```{r, eval=FALSE}
scale_y_log10()
```

### [more options]
```{r, eval=FALSE}
ggtitle("First example")
```

## Mapping aesthetics {.smaller }
All `geom_xxx()` require some *aesthetics*. `?geom_point`
<PRE style="">
> (...)
> Aesthetics
> The following aesthetics can be used with geom_point. Aesthetics are mapped 
> to variables in the data with the aes function: geom_point(aes(x = var))
> x: x position (required)
> y: y position (required)
> shape: shape of point
> colour: border colour
> size: size
> fill: internal colour
> alpha: transparency
</PRE>


Thus:
```{r, eval=FALSE}
ggplot(small)+geom_point(aes(x=carat,y=price,colour=cut))
```

also possible:
```{r, eval=FALSE}
ggplot(small,aes(x=carat,y=price,colour=cut))+geom_point()
```

```{r, eval=FALSE}
ggplot(small,aes(x=carat,y=price))+geom_point(aes(colour=cut))
```


## A subtlety to be aware of {.smaller}

Caveat: there is a difference between *setting* and *assigning* aesthetics.

```{r, echo=TRUE}
p <- ggplot(small)
```

**Assigning**
(or mapping) is done through `aes`.
```{r, echo=TRUE, fig.align='center'}
p + geom_point(aes(x = carat, y = price, colour = cut))
```

## A subtlety to be aware of {.smaller}

Caveat: there is a difference between *setting* and *assigning* aesthetics.

**Assigning**:  Done through `aes`.
```{r,eval=FALSE}
p + geom_point(aes(x = carat, y = price, colour = cut))
```

**Setting**: Fixes a parameter to a certain value and is done *outside* `aes`
```{r, echo=TRUE, fig.align='center'}
p + geom_point( aes( x  = carat, y = price), colour = "blue")
```

## A subtlety to be aware of {.smaller}

Caveat: there is a difference between *setting* and *assigning* aesthetics.

### Don't mix
But trying to set an aesthetics in `aes` produces unwanted results:
```{r, echo=TRUE, fig.align='center'}
p + geom_point( aes( x = carat, y = price, colour = "blue"))
```


## Faceting {.smaller}
`facet_xxx` are used for conditionning plots on 1 or 2 variables.

```{r, echo=TRUE, fig.align='center'}
p + geom_point(aes(x=carat,y=price))+facet_wrap(~cut)
```
```{r, eval=FALSE}
p+geom_point(aes(x=carat,y=price))+facet_wrap(~cut,nrow=1)
```
```{r, eval=FALSE}
p+geom_point(aes(x=carat,y=price))+facet_grid(cut~color)
```

## Faceting {.smaller}
`facet_xxx` are used for conditionning plots on 1 or 2 variables.

```{r, eval=FALSE}
p + geom_point(aes(x=carat,y=price))+facet_wrap(~cut)
```
```{r, echo=TRUE, fig.align='center'}
p+geom_point(aes(x=carat,y=price))+facet_wrap(~cut,nrow=1)
```
```{r, eval=FALSE}
p+geom_point(aes(x=carat,y=price))+facet_grid(cut~color)
```

## Faceting {.smaller}
`facet_xxx` are used for conditionning plots on 1 or 2 variables.

```{r, eval=FALSE}
p + geom_point(aes(x=carat,y=price))+facet_wrap(~cut)
```
```{r, eval=FALSE}
p+geom_point(aes(x=carat,y=price))+facet_wrap(~cut,nrow=1)
```
```{r, echo=TRUE, fig.align='center'}
p+geom_point(aes(x=carat,y=price))+facet_grid(cut~color)
```


## Fitting models with `geom_smooth`
`geom_smooth` is useful for displaying a trend in the data.

```{r, echo=TRUE}
p <- ggplot(small,aes(x = carat, y = price)) + geom_point()
```

By default, `geom_smooth` uses loess (akin to splines), or gam (generalised additive models) for larger datasets.
```{r, echo=TRUE, message=FALSE, fig.align='center'}
p  + geom_smooth() + facet_wrap( ~ cut)
```

## Fitting models with `geom_smooth`
You can also specify a function to use:
```{r, echo=TRUE, message=FALSE, fig.align='center'}
p  + geom_smooth( method = "lm") + facet_wrap( ~ cut)
```

## Putting everything together {.smaller}
Linear relation between `carat` and `price`, for each type of `cut`, on a log-log scale:

```{r, echo=TRUE, fig.align='center', message=FALSE}
ggplot(diamonds, aes( x = carat, y = price, colour = cut)) + 
  scale_x_log10() + scale_y_log10() + 
    geom_point( alpha = 0.3) + 
      geom_smooth(method = "lm", colour = 'black') + 
        facet_wrap( ~ cut)
```

## Many other geoms available
They all work on the same principle: mapping/setting aesthetics.
<PRE>
geom_abline  	    geom_jitter
geom_area		    geom_line
geom_bar		    geom_linerange 
geom_bin2d		    geom_path 
geom_blank		    geom_point 
geom_boxplot		geom_pointrange 
geom_contour		geom_polygon 
geom_crossbar		geom_quantile 
geom_density		geom_rect 
geom_density2d	    geom_ribbon
geom_errorbar		geom_rug 
geom_errorbarh	    geom_segment 
geom_freqpoly		geom_smooth 
geom_hex		    geom_step 
geom_histogram	    geom_text 
geom_hline		    geom_tile
geom_vline  	    etc.
</PRE>


## A couple of tricks 
- Ordering elements
- Using `aes` programmatically

## Ordering elements {.smaller} 
`wikipedia` contains the number of articles in the top 6 contributing languages. We use a bar plot to compare them. 
```{r,echo=TRUE, fig.align='center'}
wikipedia <- data.frame(
  language = c( "English", "Swedish", "German", "Dutch", "French", "Waray-Waray"), 
  articles = c(   4743614,  1959101,   1824951, 1814795,  1600497,       1259079))

ggplot(wikipedia) + 
  geom_bar(aes(x=language, y=articles), fill="#67a9cf", colour="black", stat="identity")
```

## Ordering elements {.smaller} 
`wikipedia` contains the number of articles in the top 6 contributing languages. We use a bar plot to compare them. 

```{r, echo=TRUE, fig.align='center'}
wikipedia$language <- reorder(wikipedia$language, -wikipedia$articles)

ggplot(wikipedia) + 
  geom_bar(aes(x=language, y=articles), fill="#67a9cf", colour="black", stat="identity")
```


## Using `aes` programmatically {.smaller} 
```{r, echo=TRUE, fig.align='center', fig.height=2}
variable1 <- 'carat'; variable2 <- 'price'
ggplot(small) + geom_point(aes(x = variable1, y = variable2))
```

```{r, echo=TRUE, fig.align='center', fig.height=2}
ggplot(small) + geom_point(aes_string(x = variable1, y = variable2))
```



## Good and free resources for ggplot2
- GGplot2 is vast. Not mentioned here: stats, saving as png,pdf etc., scales, themes, geographical maps etc.
- Hadley Wickham's *new* book: [https://github.com/hadley/ggplot2-book](https://github.com/hadley/ggplot2-book) 
- GGplot2 dedicated website: [http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)
- Many tutorials on the web, for example: [http://r-bloggers.com/tag/ggplot2](http://r-bloggers.com/tag/ggplot2)
- R cookbook: [http://www.cookbook-r.com](http://www.cookbook-r.com) (*e.g.* multiple plots in one figure)
- Recent 1 hour video intro to ggplot2: [https://github.com/Myfanwy/ggplot2Intro](https://github.com/Myfanwy/ggplot2Intro)
- For data manipulation in particular: `dplyr`

